Modelling palaeoecological time series using generalized additive models

### 2023-10-16 manuscript.tex
tweak to bib file
update supplementary materials to work with current gratia

\date{August 16, 2018}

@@

\usepackage{lineno}
\linenumbers

@@

\caption{Example time series; a) Small Water bulk organic matter $\delta^{15}\text{N}$ time series on a ${}^{210}\text{Pb}$ time scale, and b) Braya-Sø \uk{} time series on a calibrated ${}^{14}\text{C}$ time scale. The observations \uk{} time series have been joined by lines purely as a visual aid to highlight potential trends.}\label{fig:data-figure}
\end{figure}

@@

In \eqref{eq:gam-ef}, we further assume that the observations are drawn
from a member of the exponential family of distributions --- such as the
Poisson for count data, the binomial for presence/absence or counts from
a total --- with expected value \(\mu_t\) and possibly some additional
parameters \(\Theta\) (\(y_t \sim \text{EF}(\mu_t, \Theta)\)).
Additionally, many software implementations of the above model also
allow for distributions that are not within the exponential family but
which can be fitted using an algorithm superficially similar to the one
used to fit GAMs to members of the exponential family (e.g. Wood et al.,
2016). Common examples of such extended families include the negative
binomial distribution (for overdispersed counts) and the beta
distribution (for true proportions or other interval-bounded data).

@@

As we have already seen, polynomial basis expansions do not necessarily
lead to well-fitting models unless the true function \(f\) is itself a
polynomial. One of the primary criticisms is that polynomial basis
functions are global (Magee, 1998); the value of \(f\) at time point
\(x_t\) affects the value of \(f\) at time point \(x_{t+s}\) even if the
two time points are at opposite ends of the series. There are many other
bases we could use; here I discuss one such set of bases, that of
splines.

@@

\caption{Basis functions for the time covariate and the Small Water $\delta^{15}\text{N}$ time series. A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the rang of the data. b) shows the same CRS basis functions weighted by the estimated coefficients $\beta_j$, plus the resulting GAM trend line (black line drawn through the data). The grey points in both panels are the observed $\delta^{15}\text{N}$ values. c) A rank 7 thin plate regression spline basis for the same data.}\label{fig:basis-function-example-plot}
\end{figure}

@@

A CRS basis expansion comprised of 7 basis functions for the time
covariate in the Small Water series, is shown in Figure
\ref{fig:basis-function-example-plot}a. The tick marks on the x-axis
show the locations of the knots, which are located at the ends of the
series and evenly in between. Notice that in this particular
parametrization, the \(j\)th basis function takes a value of 1 at the
\(j\)th knot and at all other knots a value of 0.

@@

Now the i.i.d. assumption has been relaxed and a correlation matrix,
\(\boldsymbol{\Lambda}\), has been introduced that is used to model
autocorrelation in the residuals. The δ\textsuperscript{15}N values are
irregularly spaced in time and a correlation structure that can handle
the uneven spacing is needed (Pinheiro and Bates, 2000). A continuous
time first-order autoregressive process (CAR(1)) is a reasonable choice;
it is the continuous-time equivalent of the first-order autoregressive
process (AR(1)) and, simply stated, models the correlation between any
two residuals as an exponentially decreasing function of \(h\)
(\(\phi^h\)), where \(h\) is the amount of separation in time between
the residuals (Pinheiro and Bates, 2000). \(h\) may be a real valued
number in the CAR(1), which is how it can accommodate the irregular
separation of samples in time. \(\phi\) controls how quickly the
correlation between any two residuals declines as a function of their
separation in time and is an additional parameter that will be estimated
during model fitting. The model in \eqref{eq:small-gam} was fitted using
the \texttt{gamm()} function (Wood, 2004) in the \emph{mgcv} package
(Wood, 2017) for R (R Core Team, 2017).

@@

The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and
well-captures the strong pattern in the data. The trend is statistically
significant (estimated degrees of freedom = 7.95; \(F\) = 47.44,
approximate \(p\) value = \(\ll\) 0.0001). However further analysis of
the fitted model is required to answer the other questions posed earlier
about the timing of change and whether features in the trend can be
distinguished from random noise. I discuss these issues shortly.

@@

The primary issue leading to poor fit is that neither model accounts for
the different variance (known as (heteroscedasticity) of each
observation in the \uk{} record. This seemingly isn't a problem for GCV
which minimizes prediction error. The sediments in Braya-Sø are not
annually laminated and therefore the core was sliced at regular depth
intervals. Owing to compaction of older sediments and variation in
accumulation rates over time, each sediment slice represents a different
number of ``lake years''. We can think of older samples as representing
some average of many years of sediment deposition, whilst younger
samples are representative of fewer of these lake years. The average of
a larger set of numbers is estimated more precisely than the average of
a smaller set, all things equal. A direct result of this variable
averaging of lake years it that some samples are more precise and
therefore have lower variance than other samples and yet the model
assumed that the variance was constant across samples.

@@

If we want to ask whether either of the estimated trends is
statistically interesting or proceed to identify periods of significant
change, we must address the issue of uncertainty in the estimated model.
What uncertainty is associated with the trend estimates? One way to
visualise this is through a 1 - \(\alpha\) confidence interval around
the fitted trend, where \(\alpha\) is say 0.05 leading to a 95\%
interval. A 95\% interval would be drawn at
\(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
\(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
normal distribution\footnote{The 0.95 probability quantile of the
  \emph{t} distribution may be used instead, which will account for
  estimation of \(\sigma\), the variance of the data. However, given the
  number of observations, and hence residual degrees of freedom, needed
  to motivate fitting GAMs, differences between intervals computed using
  extreme quantiles of the standard normal or the \emph{t} distribution
  will be tiny.}, and \(\text{SE}(\hat{y}_t)\) is the standard error of
the estimated trend at time \(x_t\). This type of confidence interval
would normally be described as \emph{pointwise}; the coverage properties
of the interval being correct for a single point on the fitted trend,
but, if we were to consider additional points on the trend, the coverage
would logically be lower than 1 - \(\alpha\). This is the traditional
frequentist interpretation of a confidence interval. However, the GAM
described here has a Bayesian interpretation (Kimeldorf and Wahba, 1970;
Silverman, 1985; Wahba, 1983, 1990) and therefore the typical
frequentist interpretation does not apply. Nychka (1988) investigated
the properties of a confidence interval created as described above using
standard errors derived from the Bayesian posterior covariance matrix
for the estimated mode parameters. Such intervals have the interesting
property that they have good \emph{across-the-function} coverage when
considered from a frequentist perspective. This means that, when
averaged over the range of the function, the Bayesian credible intervals
shown in Figure \ref{fig:plot-fitted-models} have close to the expected
95\% coverage. However, to achieve this some parts of the function may
have more or less than 95\%-coverage. Marra and Wood (2012) recently
explained Nychka's (1988) surprising results and extended them to the
case of generalized models (non-Gaussian responses).

@@

There are a number of ways in which a simultaneous interval can be
computed. Here I follow the simulation approach described by Ruppert et
al. (2003) and present only the basic detail; a fuller description is
contained in Appendix \protect\hyperlink{appendix}{1}. The general idea
is that if we want to create an interval that contains the whole of the
true function with 1 - \(\alpha\) probability, we need to increase the
standard Bayesian credible interval by some amount. We could simulate a
large number of functions from the posterior distribution of the model
and then search for the value of \(m_{1-\alpha}\) that when multiplied
by \(\text{SE}(\hat{f}(x_t))\) yielded an interval that contained the
whole function for \((1-\alpha)\) 100\% of the functions simulated. In
practice, the simulation method of Ruppert et al. (2003) does not
involve a direct search, but yields the critical value \(m_{1-\alpha}\)
required.

@@

Derivatives of the fitted spline are not easily available analytically,
but they can be estimated using the method of finite differences. Two
values of the estimated trend, separated by a very small time-shift
(\(\Delta_t\)), are predicted from the model; the difference between the
estimated values for the two time points is an approximation of the true
first derivative of the trend. As \(\Delta_t \rightarrow 0\) the
approximation becomes increasingly accurate. In practice, the first
derivative of the fitted trend is evaluated using finite differences at
a large number of points in the time series. An approximate (1 -
\(\alpha\))100\% pointwise confidence interval can be calculated for the
derivative estimates using standard theory
(i.e.~\(\pm 1.96 \times \text{SE}(\hat{y}_t)\) for a 85\% interval) and
the covariance matrix of the spline coefficients. A (1 -
\(\alpha\))100\% simultaneous interval for the derivatives can also be
computed using the method described
\protect\hyperlink{confints}{earlier}. Periods of significant change are
identified as those time points where the (simultaneous) confidence
interval on the first derivative does not include zero.

@@

The GAM fitted to the δ\textsuperscript{15}N time series contained a
CAR(1) process to model residual temporal autocorrelation in the
residuals. The estimated magnitude of the autocorrelation is given by
the parameter \(\phi\). The estimated value of \(\phi\) for the
δ\textsuperscript{15}N series is 0.6 with 95\% confidence interval
0.28--0.85, indicating moderate to strong residual autocorrelation about
the fitted trend. The correlation function is an exponentially
decreasing function of temporal separation (\(\Delta_t\)), and whilst
observations that are a few years apart are quite strongly dependent on
one another, this dependence drops off rapidly as \(\Delta_t\) increases
and is effectively zero when samples are separated by a decade or more
(Figure~\ref{fig:car1-plot}).

@@

\caption{Estimated CAR(1) process from the GAM fitted to the Small Water $\delta^{15}\text{N}$ time series. $h(\Delta_t, \phi)$ is the correlation between residuals separated by $\Delta_t$ years, where $\hat{\phi} = \text{0.6}$. The shaded band is a 95\% pointwise confidence interval on the estimated correlation $h$.}\label{fig:car1-plot}
\end{figure}

@@

\caption{Accounting for uncertainty in age estimates whilst fitting a smooth trend to the Small Water $\delta^{15}\text{N}$ time series. (a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\text{Pb}$ inferred ages for selected depths in the sediment core (red points). The uncertainty in the ${}^{210}\text{Pb}$ inferred age is show by the red vertical bars. The fitted age model is illustrated by the solid black line. The faint grey lines are 25 random draws from the posterior distribution of the monotonically constrained GAM. The effect of age uncertainty on trend estimation is shown in b); for 100 simulations from the posterior distribution of the age model in a) a trend was estimated using a GAM with a thin plate regression spline basis and a CAR(1) process in the residuals. These trends are shown as grey lines. The combined effect of age model and fitted GAM uncertainty on the trends for the $\delta^{15}\text{N}$ time series is shown in c). The grey lines in c) are based on 50 random draws from the model posterior distribution for each of the 100 trends shown in b). For both b) and c) the black line shows the trend estimated assuming the ages of each sediment sample are known and fixed.}\label{fig:small-scam-fit-plots}
\end{figure}

@@

\hypertarget{ref-Mills2007-pq}{}
Mills, T. C. (2007). A note on trend decomposition: The ``classical''
approach revisited with an application to surface temperature trends.
\emph{Journal of applied statistics} 34, 963--972.
doi:\href{https://doi.org/10.1080/02664760701590418}{10.1080/02664760701590418}.

@@

\hypertarget{ref-R-core-team}{}
R Core Team (2017). \emph{R: A language and environment for statistical
computing}. Vienna, Austria: R Foundation for Statistical Computing
Available at: \url{https://www.R-project.org/}.

>>>>

\date{September 19, 2018}

@@

%\usepackage{lineno}
%\linenumbers

@@

\caption{Example time series; a) Small Water bulk organic matter $\delta^{15}\text{N}$ time series on a ${}^{210}\text{Pb}$ time scale, and b) Braya-Sø \uk{} time series on a calibrated ${}^{14}\text{C}$ time scale. The observations in the \uk{} time series have been joined by lines purely as a visual aid to highlight potential trends.}\label{fig:data-figure}
\end{figure}

@@

In \eqref{eq:gam-ef}, we further assume that the observations are drawn
from a member of the exponential family of distributions --- such as the
Poisson for count data, or the binomial for presence/absence or counts
from a total --- with expected value \(\mu_t\) and possibly some
additional parameters \(\Theta\)
(\(y_t \sim \text{EF}(\mu_t, \Theta)\)). Additionally, many software
implementations of the above model also allow for distributions that are
not within the exponential family but which can be fitted using an
algorithm superficially similar to the one used to fit GAMs to members
of the exponential family (e.g. Wood et al., 2016). Common examples of
such extended families include the negative binomial distribution (for
overdispersed counts) and the beta distribution (for true proportions or
other interval-bounded data).

@@

As we have already seen, polynomial basis expansions do not necessarily
lead to well-fitting models unless the true function \(f\) is itself a
polynomial. One of the primary criticisms is that polynomial basis
functions are global (Magee, 1998); the value of \(f\) at time \(x_t\)
affects the value of \(f\) at time point \(x_{t+s}\) even if the two
time points are at opposite ends of the series. There are many other
bases we could use; here I discuss one such set of bases, that of
splines.

@@

\caption{Basis functions for the time covariate and the Small Water $\delta^{15}\text{N}$ time series. A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the range of the data. b) shows the same CRS basis functions weighted by the estimated coefficients $\beta_j$, plus the resulting GAM trend line (black line drawn through the data). The grey points in both panels are the observed $\delta^{15}\text{N}$ values. c) A rank 7 thin plate regression spline basis for the same data.}\label{fig:basis-function-example-plot}
\end{figure}

@@

A CRS basis expansion comprised of 7 basis functions for the time
covariate in the Small Water series, is shown in Figure
\ref{fig:basis-function-example-plot}a. The tick marks on the x-axis
show the positions of the knots, which are located at the ends of the
series and evenly in between. Notice that in this particular
parametrization, the \(j\)th basis function takes a value of 1 at the
\(j\)th knot and at all other knots a value of 0.

@@

Now the i.i.d. assumption has been relaxed and a correlation matrix,
\(\boldsymbol{\Lambda}\), has been introduced that is used to model
autocorrelation in the residuals. The δ\textsuperscript{15}N values are
irregularly spaced in time and a correlation structure that can handle
the uneven spacing is needed (Pinheiro and Bates, 2000). A continuous
time first-order autoregressive process (CAR(1)) is a reasonable choice;
it is the continuous-time equivalent of the first-order autoregressive
process (AR(1)) and, simply stated, models the correlation between any
two residuals as an exponentially decreasing function of \(h\)
(\(\phi^h\)), where \(h\) is the amount of separation in time between
the residuals (Pinheiro and Bates, 2000). \(h\) may be a real valued
number in the CAR(1), which is how it can accommodate the irregular
separation of samples in time. \(\phi\) controls how quickly the
correlation between any two residuals declines as a function of their
separation in time and is an additional parameter that will be estimated
during model fitting. The model in \eqref{eq:small-gam} was fitted using
the \texttt{gamm()} function (Wood, 2004) in the \emph{mgcv} package
(Wood, 2017) for R (R Core Team, 2018).

@@

The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and
well-captures the strong pattern in the data. The trend is statistically
significant (effective degrees of freedom = 7.95; \(F\) = 47.44,
approximate \(p\) value = \(\ll\) 0.0001). However further analysis of
the fitted model is required to answer the other questions posed earlier
about the timing of change and whether features in the trend can be
distinguished from random noise. I discuss these issues shortly.

@@

The primary issue leading to poor fit is that neither model accounts for
the different variance (known as heteroscedasticity) of each observation
in the \uk{} record. This seemingly isn't a problem for GCV which
minimizes prediction error. The sediments in Braya-Sø are not annually
laminated and therefore the core was sliced at regular depth intervals.
Owing to compaction of older sediments and variation in accumulation
rates over time, each sediment slice represents a different number of
``lake years''. We can think of older samples as representing some
average of many years of sediment deposition, whilst younger samples are
representative of fewer of these lake years. The average of a larger set
of numbers is estimated more precisely than the average of a smaller
set, all things equal. A direct result of this variable averaging of
lake years it that some samples are more precise and therefore have
lower variance than other samples and yet the model assumed that the
variance was constant across samples.

@@

If we want to ask whether either of the estimated trends is
statistically interesting or proceed to identify periods of significant
change, we must address the issue of uncertainty in the estimated model.
What uncertainty is associated with the trend estimates? One way to
visualise this is through a 1 - \(\alpha\) confidence interval around
the fitted trend, where \(\alpha\) is say 0.05 leading to a 95\%
interval. A 95\% interval would be drawn at
\(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
\(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
normal distribution\footnote{The 0.95 probability quantile of the
  \emph{t} distribution may be used instead, which will account for
  estimation of \(\sigma\), the variance of the data. However, given the
  number of observations, and hence residual degrees of freedom, needed
  to motivate fitting GAMs, differences between intervals computed using
  extreme quantiles of the standard normal or the \emph{t} distribution
  will be tiny.}, and \(\text{SE}(\hat{y}_t)\) is the standard error of
the estimated trend at time \(x_t\). This type of confidence interval
would normally be described as \emph{pointwise}; the coverage properties
of the interval being correct for a single point on the fitted trend,
but, if we were to consider additional points on the trend, the coverage
would logically be lower than 1 - \(\alpha\). This is the traditional
frequentist interpretation of a confidence interval. However, the GAM
described here has a Bayesian interpretation (Kimeldorf and Wahba, 1970;
Silverman, 1985; Wahba, 1983, 1990) and therefore the typical
frequentist interpretation does not apply. Nychka (1988) investigated
the properties of a confidence interval created as described above using
standard errors derived from the Bayesian posterior covariance matrix
for the estimated model parameters. Such intervals have the interesting
property that they have good \emph{across-the-function} coverage when
considered from a frequentist perspective. This means that, when
averaged over the range of the function, the Bayesian credible intervals
shown in Figure \ref{fig:plot-fitted-models} have close to the expected
95\% coverage. However, to achieve this some parts of the function may
have more or less than 95\%-coverage. Marra and Wood (2012) recently
explained Nychka's (1988) surprising results and extended them to the
case of generalized models (non-Gaussian responses).

@@

There are a number of ways in which a simultaneous interval can be
computed. Here I follow the simulation approach described by Ruppert et
al. (2003) and present only the basic detail; a fuller description is
contained in Appendix \protect\hyperlink{appendix}{1}. The general idea
is that if we want to create an interval that contains the whole of the
true function with 1 - \(\alpha\) probability, we need to increase the
standard Bayesian credible interval by some amount. We could simulate a
large number of functions from the posterior distribution of the model
and then search for the value of \(m_{1-\alpha}\) that when multiplied
by \(\text{SE}(\hat{f}(x_t))\) yielded an interval that contained the
whole function for \$(1-\alpha)\$100\% of the functions simulated. In
practice, the simulation method of Ruppert et al. (2003) does not
involve a direct search, but yields the critical value \(m_{1-\alpha}\)
required.

@@

Derivatives of the fitted spline are not easily available analytically,
but they can be estimated using the method of finite differences. Two
values of the estimated trend, separated by a very small time-shift
(\(\Delta_t\)), are predicted from the model; the difference between the
estimated values for the two time points is an approximation of the true
first derivative of the trend. As \(\Delta_t \rightarrow 0\) the
approximation becomes increasingly accurate. In practice, the first
derivative of the fitted trend is evaluated using finite differences at
a large number of points in the time series. An approximate (1 -
\(\alpha\))100\% pointwise confidence interval can be calculated for the
derivative estimates using standard theory
(i.e.~\(\pm 1.96 \times \text{SE}(\hat{y}_t)\) for a 95\% interval) and
the covariance matrix of the spline coefficients. A (1 -
\(\alpha\))100\% simultaneous interval for the derivatives can also be
computed using the method described
\protect\hyperlink{confints}{earlier}. Periods of significant change are
identified as those time points where the (simultaneous) confidence
interval on the first derivative does not include zero.

@@

The GAM fitted to the δ\textsuperscript{15}N time series contained a
CAR(1) process to model residual temporal autocorrelation in the
residuals. The estimated magnitude of the autocorrelation is given by
the parameter \(\phi\). The estimated value of \(\phi\) for the
δ\textsuperscript{15}N series is 0.6 with 95\% confidence interval
0.28--0.85, indicating moderate to strong residual autocorrelation about
the fitted trend. The correlation function is an exponentially
decreasing function of temporal separation (\(h\)), and whilst
observations that are a few years apart are quite strongly dependent on
one another, this dependence drops off rapidly as \(h\) increases and is
effectively zero when samples are separated by a decade or more
(Figure~\ref{fig:car1-plot}).

@@

\caption{Estimated CAR(1) process from the GAM fitted to the Small Water $\delta^{15}\text{N}$ time series. $c(h, \phi)$ is the correlation between residuals separated by $h$ years, where $\hat{\phi} = \text{0.6}$. The shaded band is a 95\% pointwise confidence interval on the estimated correlation $c$.}\label{fig:car1-plot}
\end{figure}

@@

\caption{Accounting for uncertainty in age estimates whilst fitting a smooth trend to the Small Water $\delta^{15}\text{N}$ time series. (a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\text{Pb}$ inferred ages for selected depths in the sediment core (red points). The uncertainty in the ${}^{210}\text{Pb}$ inferred age is shown by the red vertical bars. The fitted age model is illustrated by the solid black line. The faint grey lines are 25 random draws from the posterior distribution of the monotonically constrained GAM. The effect of age uncertainty on trend estimation is shown in b); for 100 simulations from the posterior distribution of the age model in a) a trend was estimated using a GAM with a thin plate regression spline basis and a CAR(1) process in the residuals. These trends are shown as grey lines. The combined effect of age model and fitted GAM uncertainty on the trends for the $\delta^{15}\text{N}$ time series is shown in c). The grey lines in c) are based on 50 random draws from the model posterior distribution for each of the 100 trends shown in b). For both b) and c) the black line shows the trend estimated assuming the ages of each sediment sample are known and fixed.}\label{fig:small-scam-fit-plots}
\end{figure}

@@

\hypertarget{ref-Mills2007-pq}{}
Mills, T. C. (2007). A note on trend decomposition: The ``classical''
approach revisited with an application to surface temperature trends.
\emph{Journal of Applied Statistics} 34, 963--972.
doi:\href{https://doi.org/10.1080/02664760701590418}{10.1080/02664760701590418}.

@@
